Low-cost linear-scaling \emph{ab initio} molecular dynamics for weakly-interacting systems

### 2017-08-20 almo-langevin.tex
Detailed explanations on ACF

ALMO constraints restrict/prohibit/forbid electron density transfer between distant molecules but retain all other types of interaction such as long-range electrostatic, exchange, polarization, and, if the exchange-correlation (XC) functional includes it, dispersion interactions~\cite{a:theeda}. 
Since the energetic contribution of the electron transfer decays exponentially with distance, \mbox{ALMO} approximation provides a natural and accurate representation of the electronic structure of molecular systems with just [a?] few local variables if an appropriate, that is sufficiently large, localization radius is chosen. 
%The advantage of ALMO DFT is that its computational complexity is linear with respect to system size because long-range electron delocalization is ignored in both optimization steps.
%Because of the greatly reduced number of electronic variables and their robust optimization the computational cost of the implemented ALMO DFT is so low that it is [has the potential] capable of extending the range of accurate simulations to systems of thousands of molecules.
Because of the greatly reduced number of electronic descriptors and the robust optimization, the computational complexity of ALMO DFT is linear with the number of molecules while its computational overhead remains very low. These features make ALMO DFT a promising method to perform accurate AIMD simulations of large molecular systems.

@@

To adapt ALMO DFT for AIMD, the forces on atoms were computed \emph{approximately} using the idempotent DM obtained from ALMOs in a straightforward procedure that computes only the Hellmann-Feynman and Pulay forces while ignoring non-SCF forces (see the Supplemental Material). 
The difference between these forces and the exact forces obtained using perfectly-converged fully-delocalized KS orbitals is denoted $\delta f_{i\alpha}(t)$ 
%
\begin{align}
\label{eq:deltaf}
f^{\text{KS}}_{i\alpha}(t) = f^{\text{ALMO}}_{i\alpha}(t) + \delta f_{i\alpha} (t),
\end{align}
%
where $\alpha$ is a Cartesian component of the force acting on atom $i$ at time step $t$. $\delta f_{i\alpha} (t)$ comprises neglected terms that originate from %fixed terms in the trial ALMOs (i.e. orbitals obtained in the first SCF step are not re-optimized in the second step), 
a finite localization radius $R_c$ and incomplete SCF optimization of the electronic degrees of freedom. 
$\delta f_{i\alpha} (t)$ can be reduced \emph{systematically} to zero by increasing $R_c$ (Figure~\ref{fig:forcecomp} in the Supplemental Material) and, at the same time, decreasing $\epsilon_{\text{SCF}}$ -- the infinity norm of the electronic gradient of the \mbox{ALMO} energy, below which the electronic ground state is considered sufficiently close to the Born-Oppenheimer state. 
%ZZZ For $\epsilon_{\text{SCF}} \leqslant 10^{-7}$~a.u. ALMO SCF is considered converged.

\begin{figure}[t]
\includegraphics[trim={0.1cm 0cm 0.2cm 0.1cm},clip,width=8.6cm]{DeltaForceComparison_ALMO_SCF.eps}
\caption{\label{fig:forcecomp} {\color{red} All figures: $\epsilon_{\text{SCF}}$ must have ``a.u.'' units. This figure: VdWR$\rightarrow$vdWR. Hayden, you state that you average over Cartesian components of forces. Unfortunately, averaging over Cartesian components must give zero for any sufficiently large isotropic system (i.e. for every component there will be equal  opposite component). It would be better to plot the instantaneous \emph{magnitude} of $\delta \vec{f}_i(t)$, averaged over all atoms/molecules $\langle \ldots \rangle_i$. Can you make a new plot? Replace legend with: $\langle \delta f_i(t) \rangle_i$ -- red, $\langle \delta f_i(t) \rangle_{it}$ -- black, $\sqrt{\langle \delta \vec{f}_i(t) \cdot \delta \vec{f}_i(t) \rangle_{it}}$ -- green. Or remove it completely. I suggest we plot two green lines: above and below the average.} Dependence of the Euclidean norm of the instantaneous error $\langle \delta f_i(t) \rangle_i$ (red line) on the ALMO localization radius $R_c$ expressed in units of atomic van der Waals radii. Black line is the time average of the instantaneous error, green line is the standard deviation from the average. %calculated as the average difference between fully converged Born-Oppenheimer forces and fully converged ALMO DFT forces, averaged over all dimensions and all molecules for each time step. 
Forces are computed with the PBE XC functional, TZV2P basis set, and $\epsilon_{\text{SCF}} = 10^{-7}$~a.u. for the configurations from a 10-ps trajectory generated with KS-based AIMD at T=298~K.}
\end{figure}

@@

\begin{figure}
\includegraphics[trim={0.6cm 0.7cm 0.7cm 0.3cm},clip,width=8.6cm]{DeltaForceComparison_with_ACF.eps}
\caption{\label{fig:randomforce} {\color{red} The top two panels should be changed to plot the Euclidean norm of the error as described in the notes to the previous Figure. Do not forget to change the legend. For the ACF, try vector autocorrelation (see this caption below). If it is ugly, try simple scalar autocorrelation. Do not forget to subtract the non-zero mean when computing the autocorrelation. Plot (c), new x-axis label: ``Time lag $\tau$ (fs)''.} 
Red line is the Euclidean norm of the instantaneous error $\langle \delta f_i(t) \rangle_i$, black line is the time average of the instantaneous error, green line is the standard deviation from the average. Forces are computed with the PBE XC functional and TZV2P basis set for the configurations from a 10-ps trajectory generated with KS-based AIMD at T=298~K. 
% Here it is clear that Eq.\ (\ref{eq:stochastic2}) is satisfied. 
(a) $R_{c} = 1.6$ vdWR and $\epsilon_{SCF} = 10^{-2}$, 
(b) $R_{c} = 1.6$ vdWR and fully converged ALMO SCF,
(c) Autocorrelation function of the instantaneous error in panel (a) $\langle \vec{f}_i (t) \cdot \vec{f}_i(t+\tau) \rangle_t $.
}
\end{figure}

>>>>

ALMO constraints restrict/prohibit/forbid electron density transfer between distant molecules but retain all other types of interaction such as long-range electrostatic, exchange, polarization, and, if the exchange-correlation (XC) functional includes it, dispersion interactions~\cite{a:theeda}. 
Since the energetic contribution of the electron transfer decays exponentially with distance, \mbox{ALMO} approximation provides a natural and accurate representation of the electronic structure of molecular systems with just [a?] few local variables if an appropriate localization radius is chosen. 
%The advantage of ALMO DFT is that its computational complexity is linear with respect to system size because long-range electron delocalization is ignored in both optimization steps.
%Because of the greatly reduced number of electronic variables and their robust optimization the computational cost of the implemented ALMO DFT is so low that it is [has the potential] capable of extending the range of accurate simulations to systems of thousands of molecules.
Because of the greatly reduced number of electronic descriptors and the robust optimization, the computational complexity of ALMO DFT is linear with the number of molecules while its computational overhead remains very low. These features make ALMO DFT a promising method to perform accurate AIMD simulations of large molecular systems.

@@

To adapt ALMO DFT for AIMD, the forces on atoms were computed \emph{approximately} using the idempotent DM obtained from ALMOs in a straightforward procedure that computes only the Hellmann-Feynman and Pulay forces while ignoring non-SCF forces (see the Supplemental Material). 
The difference between these forces and the exact forces obtained using perfectly-converged fully-delocalized KS orbitals is denoted $\delta f_{i\alpha}(t)$ 
%
\begin{align}
\label{eq:deltaf}
f^{\text{KS}}_{i\alpha}(t) = f^{\text{ALMO}}_{i\alpha}(t) + \delta f_{i\alpha} (t),
\end{align}
%
where $\alpha$ is a Cartesian component of the force acting on atom $i$ at time $t$. $\delta f_{i\alpha} (t)$ comprises neglected terms that originate from %fixed terms in the trial ALMOs (i.e. orbitals obtained in the first SCF step are not re-optimized in the second step), 
a finite localization radius $R_c$ and incomplete SCF optimization of the electronic degrees of freedom. 
$\delta f_{i\alpha} (t)$ can be reduced \emph{systematically} to zero by increasing $R_c$ (Figure~\ref{fig:forcecomp} in the Supplemental Material) and decreasing $\epsilon_{\text{SCF}}$ -- [ZZZ the value maximum norm of the gradient of the \mbox{ALMO} DFT energy with respect to the ALMO coefficients. below which the electronic ground state is considered sufficiently close to the Born-Oppenheimer state.]
%ZZZ For $\epsilon_{\text{SCF}} \leqslant 10^{-7}$~a.u. ALMO SCF is considered converged.

\begin{figure}[t]
\includegraphics[trim={0.1cm 0cm 0.2cm 0.1cm},clip,width=8.6cm]{DeltaForceComparison_ALMO_SCF.eps}
\caption{\label{fig:forcecomp} {\color{red} All figures: $\epsilon_{\text{SCF}}$ must have ``a.u.'' units. This figure: VdWR$\rightarrow$vdWR. Hayden, you state that you average over Cartesian components of forces. Unfortunately, averaging over Cartesian components must give zero for any sufficiently large isotropic system (i.e. for every component there will be equal  opposite component). I am afraid these plots underestimate our error and mostly show finite-size effects. It would be better to plot the instantaneous \emph{magnitude} of $\delta \vec{f}_i(t)$, averaged over all atoms/molecules $\langle \ldots \rangle_i$. Can you make a new plot? Replace legend with: $\langle \delta f_i(t) \rangle_i$ -- red, $\langle \delta f_i(t) \rangle_{it}$ -- black, $\sqrt{\langle \delta \vec{f}_i(t) \cdot \delta \vec{f}_i(t) \rangle_{it}}$ -- green. Or remove it completely. I suggest we plot two green lines: above and below the average.} Dependence of the Euclidean norm of the instantaneous error $\langle \delta f_i(t) \rangle_i$ (red line) on the ALMO localization radius $R_c$ expressed in units of atomic van der Waals radii. Black line is the time average of the instantaneous error, green line is the standard deviation from the average. %calculated as the average difference between fully converged Born-Oppenheimer forces and fully converged ALMO DFT forces, averaged over all dimensions and all molecules for each time step. 
Forces are computed with the PBE XC functional, TZV2P basis set, and $\epsilon_{\text{SCF}} = 10^{-7}$~a.u. for the configurations from a 10-ps trajectory generated with KS-based AIMD at T=298~K.}
\end{figure}

@@

\begin{figure}
\includegraphics[trim={0.6cm 0.7cm 0.7cm 0.3cm},clip,width=8.6cm]{DeltaForceComparison_with_ACF.eps}
\caption{\label{fig:randomforce} {\color{red} The top two panels should be changed to plot the Euclidean norm of the error as described in the notes to the previous Figure. Do not forget to change the legend. Plot (c), new x-axis label: ``Time lag $\tau$ (fs)''. As to the ACF, one should be extremely careful how the averaging is performed. This caption below shows the ACF we need. Let me briefly explain it. The main idea is that the correct ACF could be computed from only one Cartesian component of one molecule if we could observe it for really really long time and average over many many time origins. We perform averaging over components and molecules just to improve our statistics (it is a shame to discard so many data points). We, therefore, should not compute ACF from \emph{one} time series that is already averaged over all possible components. Instead we compute and ACF for each component and then average over these ACFs. Ok, here is the procedure. Fix the time lag $\tau$. For a given time origin $t$ and molecule $i$, compute: $p_{it}(\tau)=\frac{1}{3}\delta \vec{f}_i (t)\cdot \delta \vec{f}_i (t+\tau)$, $\frac{1}{3}$ averages over components. Now, average over time origins and molecules to get $\text{ACF}(\tau) = \langle p_{it}(\tau) \rangle_{it}$. I think we do not need to subtract the mean because you already showed that it is zero. I hope there is a function in MATLAB that can do this.} 
Red line is the Euclidean norm of the instantaneous error $\langle \delta f_i(t) \rangle_i$, black line is the time average of the instantaneous error, green line is the standard deviation from the average. Forces are computed with the PBE XC functional and TZV2P basis set for the configurations from a 10-ps trajectory generated with KS-based AIMD at T=298~K. 
% Here it is clear that Eq.\ (\ref{eq:stochastic2}) is satisfied. 
(a) $R_{c} = 1.6$ vdWR and $\epsilon_{SCF} = 10^{-2}$, 
(b) $R_{c} = 1.6$ vdWR and fully converged ALMO SCF,
(c) Autocorrelation function of the instantaneous error in panel (a) $\frac{1}{3}\langle \delta \vec{f}_i (t) \cdot \delta\vec{f}_i(t+\tau) \rangle_{it} $.
}
\end{figure}
